\plaintitle

### 2018-06-20 paper.tex
Minor editing
before class commit. continue later with corrections

\section{e-Textiles}
E-textiles, the contraction of electronic textiles, are input devices fitted or integrated into the fabric of a user's clothing. These surfaces allow the user to interact with various electronic devices in his environment, as they integrates both the sensing surface, which takes the physical input like touch, grab or fingering,  and a control unit, which elaborate the input and send it to a controlled device. This means that usually an e-textile is only an input surface, which implies the necessity for the surface itself not to be rigid in order to let the skin act as a direct feedback for the user.

An example for e-Textiles is “Pinstripe” a project by the University Aachen. This technology is directly integrated in clothing by conductive threads. The threads measure folds, which are created by pinching the fabric with thumb and index finger. When the fold is rolled across the fabric the rolling gesture accompanied by the diameter of the fold acts as an input (smaller folds represent finer inputs). Those inputs can be used for example to manipulate the volume of a music player. \cite{pinstripe}%[6]

Another e-textile is the "Jogwheel", which was developed as part of the paper "Don't Mind Me Touching My Wrist[...]" \cite{touch-wrist} to investigate the social acceptance of on-body technology. The jogwheel is visually comparable to a fabric patch and is made of conductive yarn. A tactile surface structure should guide the user's fingers, which simplifies eye-free interaction. The developers describe the size of the jogwheel as twice the size of the iPod click wheel. This size should allow high precision in use and several types of one-handed gestures.
%Figures N, M
\ref{fig:1} show the jog wheel and two ways of interaction with them. The e-Textil can be placed anywhere on the user's clothing for controlling electronic devices. \cite{touch-wrist}

\begin{figure}
  \includegraphics[width=\columnwidth]{jogwheel.png}
  \caption{\emph{Jogwheel} sensing surface textile in Don't mind me touching your wrist}
  \label{fig:1}
\end{figure}
%
%
% here we can add everything to e-Textiles
Another type of e-textile is the one presented by SmartSleeve \cite{smart-sleeve}, where the fabric is an input surface made alternating conductive and insulating textile bands. Then another pressure-sensible fabric is set under the first one, beneath which there’s another fabric alternating conductive and non-conductive bands  orthogonal to the top layer. This allows to touch and register both the position and the intensity with a matrix representation. The sensor density is 1.66 sensors per square inch, which allows for a discrete sensor density. However, as the system is working with the short-circuit principle and the wires are not insulated, skin moisture can lead to false position sensing. To avoid that, the authors have put SmartSleeve over a long-sleeved running shirt. This solution, though, becomes unsuitable in summer for obvious reasons. %maybe better written
%
Project Jaquard \cite{jacquard}, instead, exploits a capacitive system where the fabric is one layer textile where some of the wires in the weft and warp are electrical wires. As the touch sensing is made with capacitance measure, this system is not vulnerable to skin moist false sensing. Different materials have been investigated to create the electrical wires, both for simplicity over the creation of the fabric and the necessity of flexibility needed for comfort. The material chosen is the copper because it has similar properties of others fabrics like cotton, silk and wool when the wire diameter is 50µm.
%
%
% At the end of this chapter we should discuss the advantages and disadvantages of the various technologies.
%
\subsection{SimpleSkin: Towards Multipurpose Smart Garments}
A related prototype created for commercially viable multi-purpose garments, the Simple Skin \cite{simpleskin} is a modular smart textile garment. It combines multiple sensing modalities, namely capacitive sensing for head and wrist movement , resistive sensing for arm movement, bio-impendance sensing for heart rate and optional smartphone interaction. This placement is intended to detect the angles of the joints to analyse posture, monitor vital signs and offer input on the socially accepted region of the lower arm with the wrist sensor. Finally, the smartphone grants the options of a regular input device or a supportive device to supply the user with  information. Its modular design offers a separation of the concerns for sensing, acquisition hardware, processing and software, allowing for the use of various combinations of different technologies on the same garment and enables modifications of the underlying fabric with detachable hardware.

\section{Advantages and disadvantages of body-worn controllers}
Body-worn controllers offer innovative solutions to a host of problems, facilitated by a combination of progressive miniaturization and sophistication of electrical components. Recent miniature controllers in the form of smart watches and wristbands can complement larger devices by supplying information or substitute them as more limited but less power-consuming alternatives. \cite{motion-ui} Although smartphones are undoubtedly the most ubiquitous form of small computing devices, improvements can be made: Mobile phones are large enough to offer extensive functionality and services, but their use can be distracting, due to the length of interactions, or even objectionable, in certain social settings. Phones are usually kept out of sight, retrieved when needed, then returned to their container or receptacle to conform to social etiquette. They also demand more concentration from the user, preventing effective use while moving or attempting to execute other tasks that require coordination. Smaller body-worn devices, for example in the form of wrist-worn contraptions, can be worn in an unremarkable and acceptable manner and remove the need for retrieval and concealment for their use. The design of new interfaces for these body-worn devices also allows for the development of more concise interaction, reducing access time to return to the previous task without interruption in a “micro-interaction” cycle. \cite{microinteraction}
Yet for all the innovative concepts developed with these technologies, they have not reached the level of ubiquity of their handheld smartphone counterparts.
This is in part due to inherent size restrictions for the sake of user convenience. Controllers are designed to be worn in a fashion conforming to conventional accessories, usually situated on the upper body for manual access. Any unnecessary bulk is trimmed to not encumber the user, especially in the case of distally attached gadgets. This comes at the cost of display sizes, resulting in diminutive screens operated by basic button controls , which can be unintuitive, or touch input, which is made difficult by “fat-finger” issues with the limited size. Operating conventional graphical interfaces like the QWERTY keyboard on very small screens is not very practical: The input method requires precious screen space to display a touchscreen keyboard and precise positioning of the fingers tips to avoid input errors. \cite{vogel7} Research efforts are being made to remedy these problems by introducing more naturalistic and expressive user interfaces for existing hardware as well as methods to influence gesture recognition by improving the goodness of input. \cite{motion-ui, vogel7} %[10, 11]

% \section{Advantages and disadvantages of E-Textiles}
% Similar to the body-worn controllers described above, E-Textiles remain niche technologies despite the range of application possibilities.
E-textiles are currently also a niche technology, despite their wide range of applications.
Just the field of pressure sensor textiles alone covers research fields that require high local resolutions, such as pressure-sore prevention, posture correction \cite{smart-rehabilitation} or motion detection due to muscular activity. \cite{vogel5}
One reason for these shortcomings lies in the overall favorability of monolithic constructions of designs and prototypes. These designs grant more tightly integrated components, resulting in more robust systems. This consequently alleviates concerns about fault tolerance, paving the way for highly performant and scalable individual components with sensor densities of multiple points per cm² in sensor areas of several m².\cite{vogel4}  A preference for these  types of systems is owed to the fact that e-textiles, more specifically smart textiles, is a relatively new field of research. \cite{simpleskin} A large part of the work achieved still focuses on addressing customization to fit the bodies of potential users or optimization to obtain useful information from user interaction. \cite{smart-rehabilitation} This approach also leads to the pursuit of more focused goals towards single-purpose gadgets that work within the confines of the more extensively researched and tested monolithic systems. Medical prototypes have notably been able to detect and distinguish simultaneous muscle activity of separate muscle groups in the same limb and are projected to accurately register more complex compound muscle activity such as movements while breathing.[16] With a focus on stability and quality of electronic components comes at the cost of extensibility and the quality of textiles. The novelty of the field  leaving less room for modification and little appeal for such products beyond experimentation, special use and utility. \cite{simpleskin, vogel4} % [12, 13]

\section{Social acceptability of touch-based Wearables}
% In this part we should discuss the issues with social acceptability
% first in general
% second regarding the specific devices we presented in the showcase
% When developing body-worn controllers, two things have to be considered. On the one hand the body part where this technology is placed - on the other, the design of the technology so that the system adapts as well as possible to the selected body part. These two parameters are a basic requirement for the most user-friendly system. [Source: Don't Mind me Touching my wrist]. However, another parameter is crucial and will be discussed in more detail in this section - social acceptability.

When developing body-worn controllers, two things have to be considered. On the one hand the body part where this technology is placed - on the other, the design of the technology so that the system adapts as well as possible to the selected body part. These two parameters are a basic requirement for the most user-friendly system. \cite{touch-wrist} However, another parameter is crucial and will be discussed in more detail in this section - social acceptability.

@@

\subsection{Testing in private (lab situation)}
% Belt (working prototype), Pinstripe, Usable Gestures, More Than Touch, SmartSleeve
Tests in a Lab offer special possibilities. On one side the subject is not distracted by other stimuli and it’s more focused on the topic he’s presented (in this case gesture testing) and on the other side researchers can verify more data recorded by the sensor, receive voice feedback by testing subject, like what they’re doing, their mental model which explains why they’re doing this gesture, while they’re performing it.
The first prototype shown is the Belt \cite{belt}, where users had a visual feedback for the belt-interaction, which was Google Glass. The main result, common to all interactions, is that the area reachable to the dominant hand is preferred. In the Belt case, in particular, the area which had a higher number of shortcuts is the front of the belt on the side of the dominant hand, where user expressed the easiness and comfortability of interacting with a body-worn device. The second case-study taken into account by this paper is the Pinstripe \cite{pinstripe} in which test subjects would perform gestures in three different context, which are sitting, standing still and walking, and  for each of them they’ll try to perform a grabbing gesture in all the sixteen areas of their body defined by the authors. For each of them they expressed the easiness and the personal acceptance of doing that gesture in that context. The most important discovery, found out also by the  papers \cite{social-comfort, belt}, is that the preferred part of the body is the upper side of the body, both for accessibility, which relates to easiness, and the comfort related to the social acceptance.
Moreover, as shown by \cite{more-touch} the mind model related to the touchscreen UI influences mostly the types of gestures (i.e. pinch to zoom is transferred from the smartphone to the textile surface)

>>>>

\section{e-Textiles}
E-textiles, the contraction of electronic textiles, are input devices fitted or integrated into the fabric of a user's clothing. These surfaces allow the user to interact with various electronic devices in his environment, as they integrates both the sensing surface, which takes the physical input like touch, grab or fingering,  and a control unit, which elaborate the input and send it to a controlled device. This means that usually an e-textile is only an input surface, which implies the necessity for the surface itself not to be rigid in order to let the skin act as a direct feedback for the user.
An example for e-Textiles is “Pinstripe” a project by the University Aachen. This technology is directly integrated in clothing by conductive threads. The threads measure folds, which are created by pinching the fabric with thumb and index finger. When the fold is rolled across the fabric the rolling gesture accompanied by the diameter of the fold acts as an input (smaller folds represent finer inputs). Those inputs can be used for example to manipulate the volume of a music player. \cite{pinstripe}%[6]
Another e-textile is the ``Jogwheel'', which was developed as part of the paper ``Don't Mind Me Touching My Wrist[...]'' \cite{touch-wrist} to investigate the social acceptance of on-body technology. The jogwheel is visually comparable to a fabric patch and is made of conductive yarn. A tactile surface structure should guide the user's fingers, which simplifies eye-free interaction. The developers describe the size of the jogwheel as twice the size of the iPod click wheel. This size should allow high precision in use and several types of one-handed gestures.
%Figures N, M
\ref{fig:1} show the jog wheel and two ways of interaction with them. The e-Textil can be placed anywhere on the user's clothing for controlling electronic devices. \cite{touch-wrist}

\begin{figure}
  \includegraphics[width=\columnwidth]{jogwheel.png}
  \caption{\emph{Jogwheel} sensing surface textile in Don't mind me touching your wrist}
  \label{fig:1}
\end{figure}
%
%
% here we can add everything to e-Textiles
Another type of e-textile is the one presented by SmartSleeve \cite{smart-sleeve}, where the fabric is an input surface made alternating conductive and insulating textile bands. Then another pressure-sensible fabric is set under the first one, beneath which there’s another fabric alternating conductive and non-conductive bands  orthogonal to the top layer. This allows to touch and register both the position and the intensity with a matrix representation. The sensor density is 1.66 sensors per square inch, which allows for a discrete sensor density. However, as the system is working with the short-circuit principle and the wires are not insulated, skin moisture can lead to false position sensing. To avoid that, the authors have put SmartSleeve over a long-sleeved running shirt. This solution, though, becomes unsuitable in summer for obvious reasons. %maybe better written
%
Project Jaquard \cite{jacquard}, instead, utilizes a capacitive system where the fabric is one layer textile where some of the wires in the weft and warp are electrical wires. As the touch sensing is made with capacitance measure, this system is not vulnerable to errors due to the skin moisture. Different materials have been investigated to create the electrical wires, both for simplicity over the creation of the fabric and the necessity of flexibility needed for comfort. The material chosen is copper because it has similar properties to cotton, silk and wool when the wire diameter is 50µm.
%
%
% At the end of this chapter we should discuss the advantages and disadvantages of the various technologies.
%
% \subsection{SimpleSkin: Towards Multipurpose Smart Garments}
The Simple Skin \cite{simpleskin} is an example of modular smart textile garment, created for commercially viable multi-purpose garments. It combines multiple sensing modalities, namely capacitive sensing for head and wrist movement , resistive sensing for arm movement \cite{vogel4}, bio-impedance sensing for heart rate and optional smartphone interaction. This placement is intended to detect the angles of the joints to analyse posture, monitor vital signs and offer input on the socially accepted region of the lower arm with the wrist sensor. Finally, the smartphone grants the options of a regular input device or a supportive device to supply the user with  information. Its modular design offers a separation of the concerns for sensing, acquisition hardware, processing and software, allowing for the use of various combinations of different technologies on the same garment and enables modifications of the underlying fabric with detachable hardware. \cite{simpleskin}

\section{Advantages and disadvantages of body-worn controllers}
Body-worn controllers offer innovative solutions to a host of problems, facilitated by a combination of progressive miniaturization and sophistication of electrical components. Recent miniature controllers in the form of smart watches and wristbands can complement larger devices by supplying information or substitute them as more limited but less power-consuming alternatives. \cite{motion-ui} Although smartphones are undoubtedly the most ubiquitous form of small computing devices, improvements can be made: Mobile phones are large enough to offer extensive functionality and services, but their use can be distracting, due to the length of interactions, or even objectionable, in certain social settings. Phones are usually kept out of sight, retrieved when needed, then returned to their container or receptacle to conform to social etiquette. They also demand more concentration from the user, preventing effective use while moving or attempting to execute other tasks that require coordination. Smaller body-worn devices, for example in the form of wrist-worn contraptions, can be worn in an unremarkable and acceptable manner and remove the need for retrieval and concealment for their use. The design of new interfaces for these body-worn devices also allows for the development of more concise interaction, reducing access time to return to the previous task without interruption in a “micro-interaction” cycle. \cite{microinteractions}
Yet for all the innovative concepts developed with these technologies, they have not reached the level of ubiquity of their handheld smartphone counterparts.
This is in part due to inherent size restrictions for the sake of user convenience. Controllers are designed to be worn in a fashion conforming to conventional accessories, usually situated on the upper body for manual access. Any unnecessary bulk is trimmed to not encumber the user, especially in the case of distally attached gadgets. This comes at the cost of display sizes, resulting in diminutive screens operated by basic button controls, which can be unintuitive, or touch input, which is made difficult by ``fat-finger'' issues with the limited size. Operating conventional graphical interfaces like the QWERTY keyboard on very small screens is not very practical: The input method requires precious screen space to display a touchscreen keyboard and precise positioning of the fingers tips to avoid input errors. \cite{vogel7} Research efforts are being made to remedy these problems by introducing more naturalistic and expressive user interfaces for existing hardware as well as methods to influence gesture recognition by improving the goodness of input. \cite{motion-ui, vogel7} %[10, 11]

Body-worn electronics in the form of e-textiles suffer the related limitations keeping them niche technologies, regardless of the range of application possibilities. Just the field of pressure sensor textiles alone covers research fields that require high local resolutions, such as pressure-sore prevention, posture correction \cite{smart-rehabilitation} or motion detection due to muscular activity. \cite{vogel5}
One reason for these shortcomings lies in the overall favorability of monolithic constructions of designs and prototypes. These designs grant more tightly integrated components, resulting in more robust systems. This consequently alleviates concerns about fault tolerance, paving the way for highly performant and scalable individual components with sensor densities of multiple points per cm² in sensor areas of several m².\cite{vogel4}  A preference for these  types of systems is owed to the fact that e-textiles, more specifically smart textiles, is a relatively new field of research. \cite{simpleskin} A large part of the work achieved still focuses on addressing customization to fit the bodies of potential users or optimization to obtain useful information from user interaction. \cite{smart-rehabilitation} This approach also leads to the pursuit of more focused goals towards single-purpose gadgets that work within the confines of the more extensively researched and tested monolithic systems. Medical prototypes have notably been able to detect and distinguish muscle activity of separate muscle groups during coactivation and are projected to accurately register more complex compound muscle activity such as movements while breathing.\cite{vogel5} Furthermore, despite the novelty of the field, modular designs rivalling the predominantly monolithic approach are being developed to improve extensibility, and increasingly maturing projects give rise to shifting the focus from stability and quality of electronic components to the quality of textiles. This development in prioritizing previously sidelined commercial concerns will create more possibilities for customization, modification and create more appeal for such products beyond experimentation and utility. \cite{simpleskin, vogel4,niche-mainstream} % [12, 13, 18]

\section{Social acceptability of touch-based Wearables}
% In this part we should discuss the issues with social acceptability
% first in general
% second regarding the specific devices we presented in the showcase
% When developing body-worn controllers, two things have to be considered. On the one hand the body part where this technology is placed - on the other, the design of the technology so that the system adapts as well as possible to the selected body part. These two parameters are a basic requirement for the most user-friendly system. [Source: Don't Mind me Touching my wrist]. However, another parameter is crucial and will be discussed in more detail in this section - social acceptability.
When developing body-worn controllers, two things have to be considered. On the one hand the body part where this technology is placed - on the other, the design of the technology so that the system adapts as well as possible to the selected body part. These two parameters are a basic requirement for the most user-friendly system. \cite{touch-wrist} However, another parameter is crucial and will be discussed in more detail in this section - social acceptability.

@@

\subsection{Testing in private (lab situation)}
% Belt (working prototype), Pinstripe, Usable Gestures, More Than Touch, SmartSleeve
Tests in a lab offer special possibilities. On one side the subject is not distracted by other stimuli and is more focused on the presented topic (in this case gesture testing). On the other side researchers can verify more data recorded by the sensor, receive voice feedback by the testing subject, like what they’re doing, their mental model which explains why they’re doing this gesture, while they’re performing it.
The first prototype shown is “Belt” \cite{belt}, here users had a visual feedback for the belt-interaction, which was Google Glass. The main result, common to all interactions, is that the area reachable to the dominant hand is preferred. In the Belt case, in particular, the area which had a higher number of shortcuts is the front of the belt on the side of the dominant hand, where users expressed the easiness and comfortability of interacting with a body-worn device. The second case-study taken into account by this paper is “Pinstripe” \cite{pinstripe} in which test subjects would perform gestures in three different contexts, which are sitting, standing still and walking, and for each of them they’ll try to perform a grabbing gesture in all the sixteen areas of their body defined by the authors. For each of them they expressed the easiness and the personal acceptance of doing that gesture in that context. The most important discovery, found out also by the papers \cite{social-comfort, belt,more-touch}, is that the preferred part of the body is the upper side of the body, both for accessibility, which relates to easiness, and the comfort related to the social acceptance.
Moreover, as shown by \cite{more-touch} the mind model related to the touchscreen UI influences mostly the types of gestures (i.e. pinch to zoom is transferred from the smartphone to the textile surface)
